---
title: 2019 COVID-19 Data Exploration
output: html_document
date: "2023-02-01"
editor_options: 
  markdown: 
    wrap: 72
---

# Analysis of COVID-19 Data from 2019

This data shows COVID-19 describes traits of individual infections of
COVID-19 recorded in 2019 around the world. This dataset can be found
[here](https://www.kaggle.com/datasets/sudalairajkumar/novel-corona-virus-2019-dataset/versions/25)

#### Import CSV and necessary packages:

```{r}
data <- read.csv("~/Desktop/COVID_R_Project/COVID19_line_list_data.csv")
```

```{r}
#import libraries
library(Hmisc)
library(dplyr)
```

#### Initial exploration of the data:

```{r}
#head(data)
```

The colnames() function shows all of the types of values in this
dataset. The describe() function shows values where there is missing
data.

```{r}
colnames(data)
```

```{r}
#see where there are potentially missing values
#describe(data) 

```

#### Adjusting for missing/erroneous data:

Most deaths are measured with 0, indicating survival, or 1, indicating
death. Adjusting for death column with death listed as date, not 1
integer:

```{r}
data$death_new <- as.integer(data$death !=0)
#if death listed as a date, corrected to death =1
unique(data$death_new) #check for only two unique values, 0 and 1 
```

### Calculating Death Rates:

With new adjusted death data, we can calculate a death rate from this
dataset is as follows:

```{r}
(sum(data$death_new)/1085) #1085 from above # of obs
```

In 2019, the global death rate was 5.8%

#### Now looking at death rate by country:

First, establishing the total case count by country:

```{r}
total_case <- data.frame(with(unique(data[c("id", "country")]), table(country)))
total_case<- total_case[order(total_case$Freq, decreasing = TRUE),]
total_case
```

Then, establishing total death count by country

```{r}
data2 <-summarise_at(group_by(data,country), vars(death_new), funs(sum(.,na.rm=TRUE)))
total_death <-data2[order(data2$death_new, decreasing = TRUE),]
total_death
```

```{r}
country_total<- merge(total_case, total_death, by = 'country')
country_total <-country_total[order(country_total$Freq, decreasing = TRUE),]
country_total
```

Is this data logical? The low death totals and rates outside of East
Asia in late 2019 are consistent with the known timeline of COVID-19
spread and awareness according to the
[CDC](https://www.cdc.gov/museum/timeline/covid19.html).

We can calculate death rates for all countries:

```{r}
country_total$death_rate<-country_total$death_new/country_total$Freq
country_total
#order based on frequency:
country_total <-country_total[order(country_total$Freq, decreasing = TRUE),]
country_total
#exclude countries with non-zero death rates:
country_total <-country_total %>%
  filter(if_all(country:death_new, ~ !.x  %in% 0))
```

Logically, it makes sense that China would have the highest case count
while the higher death rates in Iran and the Phillipines should be
looked at within the context of much lower case counts.

### Death rate related to age and gender

First we can calculate the mean age of the surviving population and not
surviving population:

```{r}
died = subset(data, death_new ==1 )
alive = subset(data, death_new ==0)
```

```{r}
#create subsets of died and alive populations
#mean(died$age) #there are missing age values in the dataset, so we need to correct for those to be ignored by R 

round(mean(died$age, na.rm = TRUE), digits = 1)
round(mean(alive$age, na.rm = TRUE), digits =1)
```

Is this statistically significant? We can assess with t-test

```{r}
t.test(died$age, alive$age, alternative = 'two.sided', conf.level = 0.95)
```

Therefore, there is a 95% chance the difference in age between the
surviving group and the died group is between 16.74 and 24.29. So our
results are statistically significant. Our p-value 2.2 e-16 \<0.05 so we
can reject the null hypothesis.

#### We can similarly assess for gender:

```{r}
men = subset(data, gender =='male' )
women = subset(data, gender =='female')
```

```{r}
#create subsets of male and female populations
#mean(died$age) #there are missing age values in the dataset, so we need to correct for those to be ignored by R 

round(mean(men$death_new, na.rm = TRUE), digits = 4) 
round(mean(women$death_new, na.rm = TRUE), digits =4)
```

So we see a death rate of 8.5% for men, and 3.7% for women. Is this
significant?

```{r}
t.test(men$death_new, women$death_new, alternative = 'two.sided', conf.level = 0.95)
```

We can say with 95% confidence that men have a 1.7% to 7.9% higher death
rate than women. Our p-value is 0.002 \<0.05 so we can reject the null
hypothesis.
